Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-11-14 manuscript/manuscript.tex
[manuscript] unify notation for notebook numbers

This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
The first notebook (00), which we will summarize in the following, showcases the entire estimation,
validation, and analysis workflow for a small example system.
The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
The seven subsequent notebooks (01--07) provide in-depth lessons on specific topics,
and the last notebook (08) contains guidelines on how to deal with common problems during MSM estimation.

@@

For the remainder of this manuscript we will walk through the first notebook (00).
In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~$0.1$~ns.
We present the results obtained in this notebook,
thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.

@@

In Markov state modeling, our objective is to model the slow dynamics of a molecular process.
In order to approximate the slow dynamics in a statistically efficient manner,
a lower dimensional representation of our simulation data is necessary.
However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint}
provide a systematic means to quantitatively compare multiple representations of the simulation data.
In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.

Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features reflecting protein structure,
we compute the (cross-validated) VAMP-2 score (notebook 00).
Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
it is important to ensure that properties that we optimize are robust as a function of lag time. 
Consequently, we compute the VAMP-2 score at several lag times (notebook 00). 
We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
We show one example of this ranking and the absolute VAMP-2 scores for lag time~$0.5$~ns in Fig.~\ref{fig:io-to-tica}b. 
We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
This suggests that backbone torsions are the best of the options evaluated for MSM construction.

@@

@@

TICA yields a representation of our molecular simulation data with a reduced dimensionality,
which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
which is shown in the showcase notebook (00).

@@

@@

@@

@@

In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA.
For the full analysis, please refer to the first notebook (00).
All notebooks as well as detailed installation instructions are available on \githubrepository{}.

@@

More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.

>>>>

This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
The first notebook~00, which we will summarize in the following, showcases the entire estimation,
validation, and analysis workflow for a small example system.
The goal of this introductory notebook~00 is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
The seven subsequent notebooks~01--07 provide in-depth lessons on specific topics,
and the last notebook~08 contains guidelines on how to deal with common problems during MSM estimation.

@@

For the remainder of this manuscript we will walk through the first notebook~00.
In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~$0.1$~ns.
We present the results obtained in this notebook,
thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
The figures that will be displayed in the following are created in the showcase notebook~00 and can be easily reproduced.

@@

In Markov state modeling, our objective is to model the slow dynamics of a molecular process.
In order to approximate the slow dynamics in a statistically efficient manner,
a lower dimensional representation of our simulation data is necessary.
However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint}
provide a systematic means to quantitatively compare multiple representations of the simulation data.
In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
In notebook~01, we present in detail how to extract features from MD datasets and how to systematically compare them.

Throughout this tutorial, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features reflecting protein structure,
we compute the (cross-validated) VAMP-2 score (notebook~00).
Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
it is important to ensure that properties that we optimize are robust as a function of lag time. 
Consequently, we compute the VAMP-2 score at several lag times (notebook~00). 
We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
We show one example of this ranking and the absolute VAMP-2 scores for lag time~$0.5$~ns in Fig.~\ref{fig:io-to-tica}b. 
We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
This suggests that backbone torsions are the best of the options evaluated for MSM construction.

@@

We demonstrate how to apply TICA, suggest how to interpret the projected coordinates, and compare the results to other dimension reduction techniques in notebook~02.

@@

TICA yields a representation of our molecular simulation data with a reduced dimensionality,
which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
which is shown in the showcase notebook~00.
A detailed comparison between different clustering techniques is provided in notebook~02.

@@

In notebook~03, we demonstrate in detail how to estimate and validate MSMs with PyEMMA.

@@

More details about (spectral) properties of MSMs and how to analyze them with PyEMMA are discussed in notebook~04 and notebook~05.

@@

In addition to a detailed demonstration of the above, notebook~06 demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.

@@

In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA.
For the full analysis, please refer to the first notebook~00.
All notebooks as well as detailed installation instructions are available on \githubrepository{}.

@@

More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
We further examine some symptoms that may indicate problematic or difficult datasets, and demonstrate how to deal with them in notebook~08.
